Overview and Summary of the First Workshop on Sustainable Software for Science: Practice and Experiences (WSSSPE1)

### 2014-03-28 WSSSPE1_report/WSSSPE1_summary_report.tex
reading and list of notes and todos
addressing Ketan's comments, and a few other small changes

\begin{abstract}
This technical report discusses the First Workshop on Sustainable
Software for Science: Practice and Experiences (WSSSPE1).  The content
is based on the workshop itself, as documented in a set of
collaborative notes taken during the workshop, including discussion
about the keynote presentations, three themed panels that focus on
papers accepted by the workshop, other discussion, cross-cutting
issues, use cases, and conclusions.  It joins a nascent literature
seeking to understand what drives software work in science and how the
reward systems of science thereby shape the type of software work
undertaken, including the extent to which developers are motivated to
build software for the long-term, for the use of others, and whether
to work collaboratively or separately. Unique perspectives were
captured about issues such as documentation of APIs, software
deployment, and development practices; licenses on software; career
paths for scientific software developers; identifiers for software
through a mechanism such as a Digital Object Identifier or
publications of ``software papers''; drawing on evidence of software
contribution, and impact, recorded in online systems such as source
code repositories like GitHub. For truly sustainable software there
should be no endpoint, as the software products continue to be used
and useful beyond the single institution, grant, and developer or
development team.

@@

Because progress in scientific research is dependent on the quality and
accessibility of software at all levels, it is now critical to address many
challenges related to the development, deployment, and maintenance of reusable
software.
%\note{Ketan: What are the levels in scientific process where software
%is required. It might be interesting to outline these stages and types of
%software required in each stage.}
In addition, it is essential that scientists,
researchers, and students are able to learn and adopt software-related skills
and methodologies. Established researchers are already acquiring some of these
skills, and in particular a specialized class of software developers is
emerging in academic environments who are an integral and embedded part of
successful research teams. The WSSSPE workshop was intended to provide a forum
for discussion of the challenges, including both positions and experiences. The
short papers and debates have been archived to provide a basis for continued
discussion, and the workshop has fed into the collaborative writing of this
journal publication \note{Ketan: we say report in abstract}. An estimate of 90 to 150 participants were present
throughout the day. Additional papers based on extended versions of workshop
submissions are expected. The level of interest in the workshop has led the
organizers, working with some of the submitters and attendees, to plan two
additional workshops (at the 2014 SciPy and SC14 conferences) and to turn the
workshop website into a community website that can be used as a focus for
further discussion and progress. Additionally, a minisymposium that aims to
further explore some of the key issues raised in WSSSPE is co-organized by a
WSSSPE1 participant at the 2014 SIAM Annual Meeting on ``Reliable
Computational Science.''

Before the WSSSPE1 workshop took place, the organizers self-published
a paper~\cite{WSSSPE1-pre-report} to document the process of
organizing and advertising the workshop, collecting and reviewing the
papers, and putting together the agenda. Section~\ref{sec:process} is
based on that publication. The remainder of this paper is based on the
workshop itself, as documented in a set of collaborative notes taken
during the workshop~\cite{WSSSPE1-google-notes}, including discussion
about the keynote presentations (\S\ref{sec:keynotes}), the three
panels (\S\ref{sec:devel}-\ref{sec:community}), other discussion
(\S\ref{sec:other}), cross-cutting issues (\S\ref{sec:cross-cutting}),
use cases (\S\ref{sec:use-cases}), and conclusions
(\S\ref{sec:conclusions}).

@@

Papers and discussion concentrated in three areas: How then \note{something missing here in this sentence --ketan} ought software  work to be visible, what are the barriers to its visibility, and what can be done to make it more visible?

@@

Katz and Proctor~\cite{Katz_WSSSPE} included a set of questions that could be
used to measure software sustainability, and though these questions might
falsely lead to yes or no answers, the complete set would determine a range of
values for sustainability. These questions are:
\begin{itemize}
\item Will the software continue to provide the same functionality in the future,
      even if the environment (other parts of the infrastructure) changes?
\item Is the functionality and usability clearly explained to new users?
\item Do users have a mechanism to ask questions and to learn about the element?
\item Will the functionality be correct in the future, even if the environment changes?
\item Does it provide the functionality that existing and future users want?
\item Does it incorporate new science/theory/tools as they develop?
\end{itemize}

@@

>>>>

\begin{abstract}
This document discusses the First Workshop on Sustainable
Software for Science: Practice and Experiences (WSSSPE1).  The content
is based on the workshop itself, as recorded in a set of
collaborative notes taken during the workshop, including discussion
about the keynote presentations, three themed panels that focus on
papers accepted by the workshop, other discussion, cross-cutting
issues, use cases, and conclusions.  It joins a nascent literature
seeking to understand what drives software work in science and how the
reward systems of science thereby shape the type of software work
undertaken, including the extent to which developers are motivated to
build software for the long-term, for the use of others, and whether
to work collaboratively or separately. Unique perspectives were
captured about issues such as documentation of APIs, software
deployment, and development practices; software licenses; career
paths for scientific software developers; identifiers for software
through a mechanism such as a Digital Object Identifier or
publication of a ``software paper''; and evidence of software
contribution and impact, recorded in online systems such as source
code repositories like GitHub. For truly sustainable software there
should be no endpoint, as the software products continue to be used
and useful beyond the single institution, grant, and developer or
development team.

@@

Because progress in scientific research is dependent on the quality and
accessibility of software at all levels, it is now critical to address many
challenges related to the development, deployment, and maintenance of reusable
software.
%\note{Ketan: What are the levels in scientific process where software
%is required. It might be interesting to outline these stages and types of
%software required in each stage.}
In addition, it is essential that scientists,
researchers, and students are able to learn and adopt software-related skills
and methodologies. Established researchers are already acquiring some of these
skills, and in particular a specialized class of software developers is
emerging in academic environments who are an integral and embedded part of
successful research teams. The WSSSPE workshop was intended to provide a forum
for discussion of the challenges, including both positions and experiences. The
short papers and debates have been archived to provide a basis for continued
discussion, and the workshop has fed into the collaborative writing of this
document. An estimate of 90 to 150 participants were present
throughout the day. Additional papers based on extended versions of workshop
submissions are expected. The level of interest in the workshop has led the
organizers, working with some of the submitters and attendees, to plan two
additional workshops (at the 2014 SciPy and SC14 conferences) and to turn the
workshop website into a community website that can be used as a focus for
further discussion and progress. Additionally, a minisymposium that aims to
further explore some of the key issues raised in WSSSPE is co-organized by a
WSSSPE1 participant at the 2014 SIAM Annual Meeting on ``Reliable
Computational Science.''

Before the WSSSPE1 workshop took place, the organizers self-published
a paper~\cite{WSSSPE1-pre-report} to document the process of
organizing and advertising the workshop, collecting and reviewing the
papers, and putting together the agenda. Section~\ref{sec:process} is
based on that publication. The remainder of this paper is based on the
workshop itself, as documented in a set of collaborative notes taken
during the workshop~\cite{WSSSPE1-google-notes}, including discussion
about the keynote presentations (\S\ref{sec:keynotes}), the three
panels (\S\ref{sec:devel}-\ref{sec:community}), 
cross-cutting issues (\S\ref{sec:cross-cutting}),
use cases (\S\ref{sec:use-cases}), and conclusions
(\S\ref{sec:conclusions}).

@@

Papers and discussion concentrated in three areas: How ought software work to be visible, what are the barriers to its visibility, and what can be done to make it more visible?

@@

Katz and Proctor~\cite{Katz_WSSSPE} included a set of questions that could be
used to measure software sustainability, and though these questions might
falsely lead to yes or no answers, the complete set would determine a range of
values for sustainability. These questions are:
\begin{itemize}
\item Will the software continue to provide the same functionality in the future,
      even if the environment (other parts of the infrastructure) changes?
\item Is the functionality and usability clearly explained to new users?
\item Do users have a mechanism to ask questions and to learn about the software?
\item Will the functionality be correct in the future, even if the environment changes?
\item Does it provide the functionality that existing and future users want?
\item Does it incorporate new science/theory/tools as they develop?
\end{itemize}

@@

% == Notes and ToDos ==
% 1. Why do we classify software cases into general purpose and special purpose? what is different? What are the sustainability challenges each face? Are they radically different? What kind of research do each category support? Are there separate questions to ask and answer for each category? Can we assume that the general purpose software has a wider scope and user community and challenges specific to these kinds of environments are not relavent in the special purpose scientific software which tend to be developed in the small?

% 2. What are the factors that has driven the development of these cases? How was the development started? What was the spark? Was there a design approach followed? Is the software modular/extensible? Can we identify unique and common factors and the reasons behind them.

% 3. Can we characterize the process itself for the selected cases: collaborative, seperate.

% 4. ToDo: summarize and report on an inclusive list (might not be exhaustive) of all the software products discussed in the workshop. (If possible color code the representaion according to a scheme based on the different dimensions pertaining to them.)

% 5. Has one or more of the following best practices played a role in the development: 
%     a) hosted community source-code and other artefacts repository
%     b) documentation, wikis and training material
%     c) mailing lists and web-presence
%     d) development process: agile, hackathons, collaborative?
%     e) regression tests, continuous integration, quality dashboard, code review

% 6. Is there a conscious effort to deal with issues such as backward compatibility, ...

% 7. Career path: are the original developer(s) still around? Is most of the discussion around career path issues a myth after all?

% 8. Is there a software paper for this software product? Cite?

% 9. Has the project proved to be sustainable so far? Does the developers, managers make conscious efforts to make sure the project is sustainable? What determines the sustainability of the project in question: science, publications, usability, funding, reusability? Perhaps all factors to some extent? What factors are dominating? Can we send out a message to current scientific project to emphasize on one or more of these or other aspects?
% ====
